\bf Adaptive Mesh Refinement for 1-D Hyperbolic PDEs

### 2015-03-06 paper_draft1.tex
paper --error estimation, references, 1d advection

\section{Overview of Paper}
\subsection{Description of AMR (Fig 2.1 from Berger-Leveque)}
An adaptive mesh refinement strategy that uses rectangular patches over Cartesian grids to refine both space and time coordinates is useful for modeling and tracking regions where the solutions are not smooth such as exhibited around shocks.

The idea behind the refinement step comes from first determining where a refinement may be useful. See section \ref{whentorefine} for details.  Examine figure \ref{fig21}.
We need to explain how to preserve conservation of flux at grid interfaces. There are essentially three cases to consider: coarse cell to coarse cell, fine cell to fine cell, and coarse cell to fine cell. Let $F_i$ be the flux at the left edge of coarse cell $i$. This is different than what we typically wrote in class, $F_{i-\frac{1}{2}}$, for convenience. Thus, updating between coarse grid cells is simply:
\begin{equation}
q_{i}^{1} = q_{i}^{0} - \frac{k}{h} (F_{i+1}^{0} - F_{i}^{0}) 
\end{equation}
\noindent where $h,k$ are the space and time lattice spacing respectively.

@@

In regions where the coarse and fine grids overlap, care must be taken. Two ghost cells are introduced, $\hat{q}_{m+1},\hat{q}_{m+2}$. The values in these 
cells are decided by interpolating the coarse values  $q_{i}^{0}, q_{i}^{1}$. At the final time (completed time step)  $q_{i}^{1}$ is replaced by the average of the 
fine grid values in the overlapped region:
\begin{equation}
q_{j-1}^{1} = \frac{1}{2} (\hat{q}_{m+1}^{2}+\hat{q}_{m+2}^{2})
\end{equation}
\noindent and the corresponding left edge flux is similarly modified so that $q_{i}^{1}$ is finally modified as:
\begin{equation}
q_{i}^{1} \rightarrow q_{i}^{1} + \frac{k}{h} ( \frac{1}{2}(\hat{F}_{m+1}^{0} - \hat{F}_{m+1}^{1}) - F_{i}^{0}) \, .
\end{equation}

\begin{figure}[h]
    \centering
    \includegraphics[width=.65\textwidth]{berg-lev-fig21}
    \caption{essential figure for adapting cartesian grids}
    \label{fig21}
\end{figure}

\subsection{Wave-propagation for 1D equations -updating grid interfaces (section 4 of paper)}
\subsection{Pseudo-code for AMR}
\subsection{Comment on determining refinement - error estimation}\label{whentorefine}
If the solution $q(x,t)$ is smooth enough, assume that our difference method $D$ has order of accuracy $s$ in both 
space and time, then the local truncation error is $q(x,t+k) -D q(x,t) = \tau(x,t) + k O(k^{s+1}+h^{s+1}))$ where $\tau$ is the leading error.
Taking two time steps suggests $q(x,t+2k) -D^2 q(x,t) = 2\tau(x,t) + k O(k^{s+1}+h^{s+1}))$.
Now, suppose we consider coarsening our grid to have space and time widths $2h$, $2k$ respectively. Let the differencing scheme be called $D_{2h}$ in this case. The local truncation error in this case reads $q(x,t+2k) -D_{2h} q(x,t) = 2^{s+1}\tau(x,t) + O(h^{s+2})$. Thus, an error growth estimate at time $t$ can be formed by taking the difference in the error present taking two steps with the regular integration scheme, and a single large step using every other grid point:

>>>>

\section{Overview of Paper}
An adaptive mesh refinement strategy that uses rectangular patches over Cartesian grids to refine both space and time coordinates is useful for modeling and tracking regions where the solutions are not smooth such as exhibited around shocks. 
\subsection{Flux updates on refined and coarse meshes}
The idea behind the refinement step comes from first determining where a refinement may be useful. See section \ref{whentorefine} for details.  
We need to explain how to preserve conservation of flux at grid interfaces, Fig \ref{fig21}. There are essentially three cases to consider: coarse cell to coarse cell, fine cell to fine cell, and coarse cell to fine cell. Let $F_i$ be the flux at the left edge of coarse cell $i$. This is different than what we typically wrote in class, $F_{i-\frac{1}{2}}$, for convenience. Thus, updating between coarse grid cells is simply:
\begin{equation}
q_{i}^{1} = q_{i}^{0} - \frac{k}{h} (F_{i+1}^{0} - F_{i}^{0}) 
\end{equation}
\noindent where $h,k$ are the space and time lattice spacing respectively.

@@

In regions where the coarse and fine grids overlap, care must be taken. Two ghost cells are introduced, $\hat{q}_{m+1},\hat{q}_{m+2}$. The values in these 
cells are decided by interpolating the coarse values  $q_{i}^{0}, q_{i}^{1}$. At the final time (completed time step)  $q_{i}^{1}$ is replaced by the average of the 
fine grid values in the overlapped region:
\begin{equation}
q_{j-1}^{1} = \frac{1}{2} (\hat{q}_{m+1}^{2}+\hat{q}_{m+2}^{2})
\end{equation}
\noindent and the corresponding left edge flux is similarly modified so that $q_{i}^{1}$ is finally modified as:
\begin{equation}\label{fluxcorrect}
q_{i}^{1} \rightarrow q_{i}^{1} + \frac{k}{h} ( \frac{1}{2}(\hat{F}_{m+1}^{0} - \hat{F}_{m+1}^{1}) - F_{i}^{0}) \, .
\end{equation}

\begin{figure}[h]
    \centering
    \includegraphics[width=.65\textwidth]{berg-lev-fig21}
    \caption{Essential figure for flux adapting Cartesian grids. The top grid is the normal coarse grid we are used to using with fluxes indicated. The middle diagram shows the case where a fine grid overlays a coarse grid and the interface region of fine and coarse grids. Recall that $2\hat{h}=h$ and $2\hat{k}=k$. The bottom diagram depicts the ghost cell region imposed on the coarse cell interface as well as the fine grid fluxes needed to bridge the interface.}
    \label{fig21}
\end{figure}

\subsection{Updating grid interfaces with wave propagation}
For conservation laws we have developed wave propagation methods in class whereby a Riemann problem is solved at each interface between grid cells. These methods can be extended beyond the linear accuracy to second order accuracy and controlled with limiters where smooth to discontinuous transitions occur. Wave propagation algorithms are written in a manner that they can be extended to hyperbolic problems not in conservation form retaining wave features but with no flux function.

With flux-differencing to ensure conservation a new coarse grid value is updated by an average of fine-grid values in any cell covered by a fine grid. The key to conservation is that flux into coarse cells is accounted for by flux out of adjacent fine cells, and vice versa.
For wave-propagation algorithms, a similar correction is needed for the waves to yield conservation.  The wave-propagation form can be used to update the values at the interface between fine and coarse grid cells on each grid independently using Òghost-cellÓ values as needed near grid interfaces.

Whether conservative or non-conservative systems are treated, second-order corrections are still written as flux-differences and utilize the flux correction as eqn(\ref{fluxcorrect}). 
However, the first-order upwind terms written in terms of the fluctuations must be handled differently.

To maintain conservation we must {\it first} solve the Riemann problem between the ghost cell value $\hat{q}_{m+1}^{0}$ on the fine grid and the coarse grid value $q_{j}^{0}$, and add to the coarse cell value $q_{j}^{1}$ the resulting total fluctuation $A^{-}\Delta q + A^{+}\Delta q$ weighted by $\hat{k}/h = 1/2$ since the time step is $\hat{k}$ while the cell size is $h$. 
{\it Second} we must also solve a Riemann problem on the fine grid between $\hat{q}_{m+1}^{1}$ and $q_{j}^{0}$
and add these fluctuations into $q_{j}^{1}$.

The correction step extends to an arbitrary hyperbolic system where we use a Riemann solver to obtain $A^{-}\Delta q$ and  $A^{+}\Delta q$ from the two states $\hat{q}_{m+1}^{0}$ and $q_{j}^{0}$.  This step is made in each of the $R$ time steps on the refined grid within the single coarse-grid step, where $R$ is the refinement ratio ($2X$ in our examples so far). \\

\noindent The fix-up algorithm:
\begin{description}
  \item[for $N = 0 , R-1$] \hfill \\
  solve the Riemann problem with data $\hat{q}_{m+1}^{N}$ and $q_{j}^{0}$ to compute $A^{-}\Delta q$ and  $A^{+}\Delta q$\\ \hfill
  update $q_{j}^{1} \rightarrow q_{j}^{1} + \frac{\hat{k}}{h}$ ($A^{-}\Delta q + A^{+}\Delta q$)
\end{description}

\subsection{Pseudo-code for AMR}
\subsection{Comment on determining refinement - error estimation}\label{whentorefine}
If the solution $q(x,t)$ is smooth enough, assume that our difference method $D$ has order of accuracy $s$ in both 
space and time, then the local truncation error is $q(x,t+k) -D q(x,t) = \tau(x,t) + k O(k^{s+1}+h^{s+1}))$ where $\tau$ is the leading error.
Taking two time steps suggests $q(x,t+2k) -D^2 q(x,t) = 2\tau(x,t) + k O(k^{s+1}+h^{s+1}))$.
Now, suppose we consider coarsening our grid to have space and time widths $2h$, $2k$ respectively. Let the differencing scheme be called $D_{2h}$ in this case. The local truncation error in this case reads $q(x,t+2k) -D_{2h} q(x,t) = 2^{s+1}\tau(x,t) + O(h^{s+2})$. Thus, an error growth estimate at time $t$ can be formed by taking the difference in the error present taking two steps with the regular integration scheme, and a single large step using every other grid point:

A similar modification must be made in each of the R time steps on the refined grid within the single coarse-grid step, where R is again the refinement ratio. The fix-up algorithm thus takes the form
?for N = 0, 1, . . . , R ? 1 do
solve the Riemann problem with data qö and q
to compute A?Æq and A+Æq, update q1 := q1 + kö (A?Æq + A+Æq).
jjh
For the case of a conservation law, this will restore conservation, and, in fact, agrees
with the flux function modification (2.4) if numerical fluxes are defined by (3.9).
Note that, in view of (3.8), in the conservation law case the coarse-grid value qj1 is
N0
simply updated by the entire flux difference f (qö ) ? f (q ), and we wouldnÕt need
m+1 j
to actually solve the Riemann problem. However, implementing it as presented above
gives a uniform and general formulation.
